### 2022-05-27 doc/VerticalArrangements.tex
incorproated peggy's edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory}
We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2021}.  This framework allows us to study a variety of merger configurations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Empirical Application (In Progress)}

@@

The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities. Haulers collect municipal solid waste (MSW) from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills. Waste disposal (upstream) is a required input into waste collection services (downstream). Some haulers are vertically integrated and operate their own disposal facilities. Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity. Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens’ waste themselves. Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal. Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals’ cost of disposal in order to benefit its own collection operations. Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business and their intensity of hauling competition with prospective disposal customers.

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{\tiny{Competitive Impact Statement: U.S. and Plantiff States v. Waste Management, Inc. and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{\tiny{Waste 360, \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{\tiny{Competitive Impcat Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{\tiny{Waste Business Journal, \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}} In the DOJ complaint filed in the Republic-Santek case, vertical competitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{\tiny{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}}

Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements. The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.

@@

The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers. The CIS notes that Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors. In MSW disposal, the CIS identifies only one other significant competitor and Republic and Santek as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere. Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.

In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, which was the parties' sole competitor in the MSW disposal market. Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated. Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consumation since the merger was completed before the filing of the CIS for the Republic-Santek merger. The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.

@@

Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) on Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019. This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee. The MSW disposal market definition follows that outlined in the DOJ CIS and attributing share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only). Thus, the total market quantity is defined as all MSW volumes originating in Hamilton county.

Disposal volumes are measured in tons and have been combined across landfills owned by the same firm. The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three market participants each with less than 0.5\% share receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined. Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis. After rescaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.

@@

However, the market shares required for implementation of the Sheu and Targain (\textbf{add ref}) framework are expressed as upstream-downstream market participant pairs. Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share. Capacity constrained, integrated firms (Republic disposal-Republic collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal-Republic collection, Waste Connections disposal-Republic collection). Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.

@@

%\subsubsection{Chattanooga Area Results}
%In progress.

@@

@@

\begin{figure}
\centering
\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
\caption{Republic/Santek Simulation Results}
\label{fig:TrashSimsFirms}
\end{figure}
\end{document}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Theory\label{sec:theory}}
We begin by describing our basic framework, which uses a downstream Bertrand logit model embedded in an upstream Nash Bargaining setup, taken from \citeN{ST2021}.  This framework allows us to study a variety of merger configurations.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Empirical Application}

@@

The vertical supply chain in the solid waste industry is primarily comprised of waste collection operations or ``haulers'' and waste disposal facilities. Haulers collect municipal solid waste (MSW) from businesses and residences and must dispose of it at a lawful disposal site, predominantly landfills. Waste disposal (upstream) is a required input into waste collection services (downstream). Some haulers are vertically integrated and operate their own disposal facilities. Vertically-integrated haulers typically prefer to dispose of waste at their own disposal facilities and may also sell a portion of their disposal capacity. Disposal customers include private waste haulers without their own disposal assets (“independent haulers”) as well as local governments that collect their citizens' waste themselves. Due to strict laws and regulations that govern the disposal of MSW, there are no reasonable substitutes for MSW disposal. Thus, mergers that combine hauling and disposal assets may incentivize the merged entity to raise its hauling rivals' cost of disposal in order to benefit its own collection operations. Whether or not the merged entity is both able and incentivized to undertake such action depends upon the extent of its market power in the local disposal market, the merging parties’ profit margins in each line of business, and their intensity of hauling competition with prospective disposal customers.

In 2020, Waste Management, the largest waste management company in the U.S., acquired Advanced Disposal Services (ADS), previously the fourth largest company, for \$4.6 billion.\footnote{Competitive Impact Statement: U.S. and Plantiff States v. Waste Management, Inc. and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1330596/download}} GFL Environmental also acquired WCA Waste Corporation for \$1.2 billion in 2020.\footnote{Waste 360,\tiny{ \url{https://www.waste360.com/business/breaking-gfl-acquires-wca-waste-corp-121-billion}}} Republic Services, the second largest waste management company in the U.S., acquired Santek Environmental in 2021.\footnote{Competitive Impcat Statement: U.S. and State of Alabama v. Republic Services, Inc.and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382626/download}}  These three merged companies along with Waste Connections, the third largest waste management company in the U.S., are estimated to control over 60\% of available landfill capacity nationally and also rank among the top haulers nationwide.\footnote{Waste Business Journal,\tiny{ \url{https://www.wastedive.com/news/public-companies-increased-control-of-74b-us-waste-industry-in-2018/556079/ }}} Concentration in local markets varies substantially, however, in both the upstream and downstream markets.

%Concerns about vertical competitive effects were raised in both the Republic-Santek and the Waste Management-ADS transactions. Both the Solid Waste Agency of Lake County, IL and the Solid Waste Agency of Northern Cook County submitted comments in opposition to the proposed asset divestiture from Waste Management-ADS, stating that a vertically integrated competitor was needed to maintain competition in their local market post-merger.\footnote{\tiny{See  Comments by SWALCO and SWANCC : U.S. and Plaintiff States v. Waste Management, Inc., and Advanced Disposal Services, Inc., \url{https://www.justice.gov/atr/case-document/file/1377646/download}.}}
For example, in the DOJ complaint filed in the Republic-Santek case, horizontal anticompetitive effects were alleged for four SCCW collection markets and two MSW disposal markets. In addition, vertical anticompetitive effects were alleged to arise from the combination of their integrated assets in the Chattanooga area.\footnote{See U.S. and State of Alabama v. Republic Services, Inc. and Santek Waste Services, LLC, \url{https://www.justice.gov/atr/case-document/file/1382031/download}}
%Our application in progress further identifies local markets in which these acquisitions result in: 1) only horizontal combinations of assets, 2) only vertical combinations of assets, and 3) combinations of vertically integrated assets in the presence of existing integrated competitors to analyze the welfare effects from mergers with complex vertical arrangements. The next sections present preliminary results on the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.
The next sections present results of applying the model of Section \ref{sec:theory} to the merger of vertically integrated assets in the presence of other integrated and unintegrated rivals in the context of Republic and Santek's merger in the Chattanooga area.

@@

The Competitive Impact Statement (CIS) filed by the DOJ in association with the Republic-Santek merger describes the alleged lost competition in the ``Chattanooga, Tennessee and North Georgia area'', subsequently referred to as the Chattanooga Area, due to lost horizontal competition in MSW disposal and small-container commerical waste (SCCW) collection as well as raising rivals costs in the SCCW collection market by raising the MSW disposal costs of independent haulers. The CIS notes that pre-merger Republic and Santek combined served approximately 73 percent of the SCCW collection market with three other significiant competitors. In MSW disposal, the CIS identifies only one other significant competitor pre-merger and Republic and Santek combined as serving approximately 82 percent of the market, disposed of either directly in the merging parties’ landfills within the area or passing through their transfer stations in Chattanooga before ultimately being disposed of in the parties’ landfills elsewhere. Thus, pre-merger both parties were large, vertically integrated competitiors in the Chattanooga Area.

In addition, another large, vertically integrated waste company existed in the market at the time of the merger, Waste Connections, and was the parties' sole competitor in the MSW disposal market. Waste Management and ADS owned collection assets in the area but were not vertically integrated in this market, as demonstrated by MSW disposal data discussed in Section{\ref{chatt_disposal_vol}}.\footnote{Firms that are national competitors and vertically integrated in other markets are known to enter contracts with each other to dispose of waste on advantageous terms that may make them effectively vertically integrated. Ignoring these contracting relationships may underestimate the number of effectively vertically integrated competitors in the market.} We treat Waste Management and ADS as a single, merged entity regardless of whether the data source predates the merger consumation since the merger was completed before the filing of the CIS for the Republic-Santek merger. The final significant participant in the SCCW collection market is a major regional firm that is not identified by name in the CIS.

@@

Data on MSW disposal comes from data collected by TDEQ (Tennessee Department of Environmental Quality) consisting of Class 1 landfill ownership, Class 1 Solid Waste Origin Reports, and waste disposal by county for 2019. This data identifies the origin county and destination landfill, including owner and operator information, for all MSW produced in Tennessee as well as waste volumes passing through transfer stations in Tennessee. The MSW disposal market definition follows that outlined in the DOJ CIS and attributes share to the company owning the final disposal landfill (i.e., ignoring transfer stations which are an intermediate disposal site only). Thus, the total market quantity is defined as all MSW volumes originating in Hamilton County, Tennessee where Chattanooga is located.

Disposal volumes are measured in tons and have been combined across landfills owned by the same firm. The TDEQ data identifies landfills owned or operated by Republic, Santek, Waste Connections, and three other market participants (each with less than 0.5\% share) receiving MSW volumes originating in Hamilton County.\footnote{The City of Chattanooga and Marion County landfills are municipally owned and operated, accounting for less than 1\% combined. Global Envirotech is a privately-owned transfer station that sends its 0.03\% share to an out of state landfill in GA.} These three fringe participants have been excluded from the analysis. After rescaling, the resulting market shares are Republic, 28.4\%, Santek, 54\%, and Waste Connections, 17.6\%.

@@

However, the market shares required for implementation of the Sheu-Targain (\textit{\textbf{add ref}}) framework are expressed as upstream-downstream market participant pairs. Following the discussion of supply relationships in Section \ref{chatt_disposal_vol}, vertically-integrated pairs without capacity constraints (i.e., Santek disposal-Santek collection, Waste Connection disposal-Waste Connections collection) are assigned their full collection share. Capacity constrained, integrated firms (Republic disposal-Republic collection) are assigned their collection share up to their available capacity with the remainder allocated by residual disposal share to pairs with the respective upstream firms (i.e., Santek disposal-Republic collection, Waste Connections disposal-Republic collection). Unintegrated firms' collection shares are  distributed among the upstream suppliers with available capacity by residual share as well.

@@

\subsubsection{Chattanooga Area Merger Simulation Results}

Overall, the merger between Republic and Santek is estimated to lower upstream prices by 2.8\% and increase downstream prices by 12\%, resulting in \$16 million of harm to consumers and total producer benefits of approximately \$13.4 million. Thus, on net the model predicts the merger would be harmful despite the presence of significant efficiencies from EDM that reduce prices upstream. These aggregate effects obscure the complicated competitive effects between these interdependent firms though. These effects are illustrated in more detail in Figure \ref{fig:TrashSimsFirms}.

Disposal prices for pre-merger integrated pairs are unchanged post-merger. However, MSW disposal prices increase to all rival haulers post-merger (i.e., all unitegrated pairs). The merged firms' disposal prices to downstream rivals are estimated to increase 88\% post-merger, and Waste Connections prices to its downstream rivals are estimated to increase 22\%. For the merged firm the price increase is driven by both the effects of increased bargaining leverage and upward pricing pressure, whereas, Waste Connections' price increase is driven only by the change in bargaining leverage. These large price increases are offset in the aggregate, volume-weighted estimate by the large decrease in the disposal price for the integrating pair post-merger due to the EDM effect, and the cost to Republic for volume disposed with Santek decreases by 60\%.

Turning to the downstream market, all collection prices increase post-merger. The merging parties' post-merger collection prices increase by approximately 20\% and share decreases by about 8 percentage points overall. The pre-merger integrated Republic pair's price increases 21\% from the effect of upward pricing pressure alone and Santek's increases by 23\% from a combination of UPP and increased bargaining leverage. The newly integrated Santek-Republic pair's price is estimated to increase by 9\%, demonstating that the upstream efficiencies are not fully passed-through to consumers. Pre-merger integrated rival, Waste Connections, collection price increases by approximately 5\% and share increases by 6 percentage points. Unintegrated rivals' collection prices each increase by about 9\%, which again demonstrates incomplete pass-through of the change in upstream costs, and share increases by 2 percentage points each.

\subsubsection{Merger Simulation of Collection Market Only}

We compare these results to those from a merger simulation considering only a merger in the collection market under logit demand, treating each upstream-downstream pair as if it were an independent firm, as shown in Figure \ref{fig:TrashSimsCompare}. Ignoring these vertical relationships results in an estimated 9.4\% increase in collection prices with a 15\% increase in the merging parties' price, consumer harm of \$9.6 million and a producer benefit of \$7.1 million. While net harm is nearly the same under both models at about \$2.5 million, estimated consumer harm is 67\% higher in the model including vertical relationships and is offset by markedly larger producer benefits. Further, the collection-only model overestimates diversion to the unintegrated rivals and as a result underestimates the market price effect due to UPP from the merger. (See also Table \ref{tab:trash_merged}.)

The Republic-Santek merger was ultimately settled with the DOJ through a divestiture. According to the final judgement, in the Chattanooga Area the parties were required to divest Santek's SCCW collection assets  as well as two landfills and a transfer station in the Chattanooga area.\footnote{Final Judgment: U.S and Plaintiff States v. Republic Services, Inc. and Santek, LLC, \url{https://www.justice.gov/atr/case-document/file/1408616/download}} Our results suggest that a divestiture in the collection market alone likely would not have sufficiently remedied the anticompetitive effects of the merger.

@@

\begin{table}
\centering
\small{\include{trash_merged_output}}
\caption{Predicted Market Shares for Integrated Model vs. Collection Only Model}
\label{tab:trash_merged}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[scale=1.0]{output/TrashSimsFirm.png}
\caption{Republic/Santek Simulation Results}
\label{fig:TrashSimsFirms}
\end{figure}

\begin{figure}
\centering
\includegraphics{output/TrashSimsCompare.png}
\caption{Merger Simulation Results Comparison}
\label{fig:TrashSimsCompare}
\end{figure}

\end{document}
